Supplementary material for "XXX Title"

### 2017-03-13 article_genome_scan.tex
update

To emphasize the utility of RDA, we compared to pcadapt from which the idea of using multivariate method for genome scan is based. On the simulated dataset, we retain $K=5$ axis to compute Mahalanobis distances  as it seems to explain the main amount of variance in the data using scatter plots. To control for false positive, we used the same qvalue threshold (i.e. <= 10\%).

@@

	A lattice of 8x8 populations is simulated (i.e. 64 populations in total). Each population is initialized with 200 diploid individual with random genotypes. Migration is set to 0.1 so that population structure must be very smooth and genetic differentiation must show an isolation by distance pattern over the 64 populations. This is where pcadapt is best designed for. Loci are biallelic (0 or 1) like SNPs. Allele frequency of the whole population is initialized at 0.5. 1000 loci are defined. They are separated in 200 chuncks of 5 SNPs in physical linkage with recombination rate between adjacent loci fixed at 0.1. 3 different Traits are coded by a group of 10 different loci. The first trait is coded by loci 1, 11, 21, ..., 91. The trait value is simply the sum of genotype value and therefore can take value between 0 an 20. For the sake of realism, we add to each trait a random noise (non heritable variation) drawn from a normal distribution $N(0,2)$. The second trait is coded by loci 101, 111, ..., 191 and the third is coded by loci 201, 211, ..., 291. Each trait is therefore coded by free recombining SNP loci. In other words, there are 30 coding SNPs among 1000. Selection can have an effect on linked loci, for instance, loci 2, 3, 4 and 5 can be impacted by selection on locus 1. However, recombination is high enough (0.1) to expect a limited linkage effect. We have defined 10 different environmental variables. The first one determines the selective pressure on trait 1, the second one on trait 2 and the third one on trait 3. The first environment variable is a quadratic gradient coded by function  $env1 = -(\cos(\theta)*(i-3.5))^2 -(sin(\theta)*(j-3.5))^2 + 18, \theta = \pi/2$, i and j being the population indicator on the 8x8 lattice. The second one is a linear plan gradient coded by function $env2 = h*\cos(\theta)*(i-1) + h*\sin(\theta)*(j-1) + k$ with $h=2$, $\theta = \pi/4$ and $k=3$. The third environment variable simulates a coarse environment with value $env3 = 2$ for all populations except population (i,j) = {(2,2), (2,3), (3,2), (3,3), (6,2), (6,3), (7,2), (7,3), (2,6), (2,7), (3,6), (3,7), (6,6), (6,7), (7,6), (7,7)} for which $env3 = 18$. Env4, env5 and env6 have exactly the same equation than env1, env2 and env3 respectively. The remaining 4 environment variable are similar to env2 but with different value of h and $\theta$. Env7 has $h=2$, $\theta = 0$ and $k=3$. Env8 has $h=2$, $\theta = \pi/4$ and $k=0$. Env9 has $h=1$, $\theta = \pi/4$ and $k=4$. Env10 has $h=0.5$, $\theta = \pi/4$ and $k=8$. Graphical representation of mean environmental value for environment 1, 2 and 3 is given in Fig. \ref{simulatedenvir}. Environment 4, 5 and 6 have respectively the same mean value spatial distribution. For a graphical representation of environment 7 to 10, see supplementary material. Environmental equation gives a mean value of the environment variable. To avoid colinearity between environments variable, we added noise by drawing an environment value within a normal distribution $N(\mu=env, \sigma=1)$. Fitness for each trait is set to be $-e^{((x-env)^2/(2*\omega^2))}$, $x$ being the quantitative trait value, $env$ the environmental value and $\omega$ is defining selection strength and has been set to 10 which in our experience seems sufficient for loci to be often detected. To get the overall fitness for a given individual, fitness associated to each trait are multiplied. Fitness are relative and selection arises on parents and determine their number of offsprings. Simulations are made across 500 generations. At the end of simulation, we sample 10 individuals per population. Therefore, we have a sample of 640 individuals with 1000 SNP-like loci.

\subsection{Real dataset}
The Loblolly pine dataset is a sample of 682 individuals genotyped on 1,730 SNPs selected in ESTs \citep{Eckert2010}. 60 climatic variables were available and summarized by the authors in the five first axis of a PCA. The first axis, PC1 is mainly linked to latitude, longitude, temperature, and winter aridity. PC2 is linked to longitude, spring-fall aridity, and precipitation. We inputed the missing data using a very simple algorithm implement in function sing.im of the R package linkim \citep{Lachenbruch2011}. It imputes the missing value based on the observed data proportions.  We used $K=4$ axis to compute Malahanobis distances.

The Chinook salmon consists of 19 703 SNP loci genotyped on 1956 total individuals pooled in 46 collections. \citet{Hecht2015} have estimated that between 5.8 and 21.8\% of genomic variation can be accounted for by environmental features, and 566 putatively adaptive loci were identified as targets of environmental adaptation. Therefore this dataset is a good candidate to test ACP and RDA approaches to detect outliers and selective gradients. Five variables (MigDistKM, StreamOrder, bio03, bio17 and bio18) have been used among  24 different climate and environmental variables because they have been tested as significantly associated with the SNP variation rangewide \citep{Hecht2015}. MigDistKM stands for Migration distance from collection site to ocean (km),
StreamOrder for Stream Order of collection site using Strahler method, bio03 for Isothermality, bio17 for Precipitation of Driest Quarter (mm) and bio18 for Precipitation of Warmest Quarter (mm). We could have tested more variable but this is just an illustration and is by no mean an extensive study of this species. Since data are pooled, we have randomly created a sample of 100 individuals for each collection based on the allele frequencies to be able to analyze the data following our individual based pipeline. We used $K=4$ axis to compute Malahanobis distances.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[height=0.21\textheight]{figures/sim5_pcadapt.eps}
\end{center}
\caption{Manhattan plot of the result of pcadapt on a simulated data set.}%
\label{fig:pcadapt}%
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[height=0.21\textheight]{figures/sim5_capscale.eps}
\end{center}
\caption{Manhattan plot of the result of genome scan using RDA on a simulated data set.}%
\label{fig:rda}%
\end{figure}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[height=0.4\textheight]{figures/rdaA.eps}
\end{center}
\caption{RDA on the adaptively enriched genetic space. We discarded the individual points for readability. Dots represents outliers SNPs.  $R^2$ of envir1 with the first, second and third axis is (0.02\%, 77.5\%, 14.5\%), envir2 is (99.3\%, 0.003\%, 0001\%) and envir3 is (0.009\%, 0.82\%, 64.7\%)}%
\label{fig:rdaA}%
\end{figure*}

@@

\subsection{Loblolly Pine}

\subsection{Chinook Salmon}

Our analysis of Chinook Salmon gave a list of 27 SNPs (Tab. \ref{tab:salmon}). From the material of \citet{Hecht2015}, we extracted their matching with coding sequences and the associated annotation.

@@

Fig \ref{fig:pcadapt} shows that pcadapt approach works well when the environmental gradient and the selective pressures are acting in the same direction than the geographical pattern of isolation by distance. Whereas when the environmental gradient is quadratic on the geographical range (QTL1) or when it is a coarse environment (QTL3). Indeed, we can hypothesize that the PCA ordination fails at orienting the genetic space differentiation into the direction of environment 1 and environment 3 therefore leaving no chance to detect any outliers on the QTL influenced by these environmental variables. Fig \ref{fig:rda} shows that RDA has a much better behavior than pcadapt by taking advantage of using informations of environmental local conditions.

Both methods have a good control of false discovery rate ($8.36 \times 10^{-2}$ for pcadapt and $8.51 \times 10^{-2}$ for RDA). Results summarized on Fig \ref{fig:performance} is confirming that overall RDA shows better performance at detecting true outliers since it succeeds to detect quite often QTL1 and QTL3 SNPs. It seems however less efficient at detecting QTL3 outliers but this might be due to the fact that local adaptation on a coarse environment is more difficult  that adaptation on a smooth environmental gradient as environment 1 and 2. These simulations plead in favor of using constrained ordination method instead of PCA when non genetic data such as environmental variable are available in order to orientate the axis in the direction of informative gradients.

@@

From the analysis of Chinook Salmon, we picked up some genes that can be interpreted regarding to the environmental variable. For instance, a heat shock protein which are known to be involved in adaptation to temperature or lipolysis-stimulated lipoprotein receptor which are involved in regulation of lipid metabolic process. This latter process can reasonably thought to be involved in adaptation to food abundance and the need for salmon to migrate on a short or long distance.
Outliers picked up by RDA based method are more easily interpretable as this is illustrated by Fig SXXX. This was of course expected since the ordination is constrained by environmental axis. One can identify an outlier to a selective gradient which was used in the inference process whereas outliers picked up by PCA based method have to be interpreted a posteriori and this can be more difficult to justify. It is however noticeable that a substantial proportion of outliers are shared between PCA and RDA based methods which highlight the fact that both approach are very similar. The lack of interpretability of PCA method is compensated by the fact that it is a blind method in regard to environmental data so it can pick up genes that we will miss using RDA because we miss some crucial environmental data. 
In conclusion, this paper shows that both methods are complementary and should be used simultaneously on the same dataset. RDA will be more easily interpretable in terms of mechanism and should pick up more genes when some important selective gradient are identified but PCA might be able to pick up outliers where no environmental data are available. However this latter method will miss outliers when selective gradient are poorly correlated to the population structure (i.e. geographical distance between populations and individuals). It is therefore crucial when performing landscape genomics to clearly characterize the environmental condition by measuring important variable that are suspected to put selective pressures on the species of interest. This can be done through RDA as it has been validated in our simulations and in the real dataset on Chinook Salmon. 
A myriad of constrained ordination method exist that could be used and allows to control for confounding variable (i.e. partial RDA) for instance altitude, latitude and longitude or allow to perform none linear regression between genetic and environmental data (i.e. LVM - Latent Variable Model). This is still to be tested and explored using both simulation and real data set validation.

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Mon Jan 30 16:49:00 2017
\begin{table}[ht]
\centering
\caption{List of SNPs with $qvalue < 0.1$ and their matching with coding sequence when available.}%
\label{tab:salmon}%
\begin{tabular}{rll}
  \hline
 & Locus & Sequence Description \\ 
  \hline
1 & 8760\_60 & cell migration-inducing and hyaluronan-binding partial \\ 
  2 & 11727\_44 & protein argonaute-1 \\ 
  3 & 15784\_70 & dna polymerase epsilon subunit 4 \\ 
  4 & 19372\_14 & protein fam122a-like isoform x1 \\ 
  5 & 19510\_54 & pantothenate kinase mitochondrial-like \\ 
  6 & 19809\_36 & eukaryotic translation initiation factor 3 subunit j \\ 
  7 & 22558\_48 & zinc finger protein gfi-1b-like \\ 
  8 & 29912\_62 & g protein-activated inward rectifier potassium channel 1-like isoform x1 \\ 
  9 & 30253\_61 & solute carrier family 1 (glial high affinity glutamate transporter) member 3b isoform x1 \\ 
  10 & 30495\_21 & c-jun-amino-terminal kinase-interacting protein 4 \\ 
  11 & 33486\_16 & heat shock 70 kda protein 12a isoform x3 \\ 
  12 & 39480\_19 & ras association domain-containing protein 4 \\ 
  13 & 40284\_30 & rna-binding single-stranded-interacting protein 2-like isoform x3 \\ 
  14 & 41648\_34 & afadin- and alpha-actinin-binding protein \\ 
  15 & 46982\_22 & unnamed protein product \\ 
  16 & 50054\_21 & protein fam92a1-like isoform x1 \\ 
  17 & 54261\_58 & leukotriene b4 receptor 1-like \\ 
  18 & 54497\_54 & ankyrin repeat domain-containing protein 50-like \\ 
  19 & 56375\_14 & mms19 nucleotide excision repair protein homolog \\ 
  20 & 60067\_64 & e3 ubiquitin-protein ligase trim37-like \\ 
  21 & 66930\_15 & tubulin polyglutamylase ttll4-like isoform x2 \\ 
  22 & 69650\_61 & monocyte to macrophage differentiation factor 2 \\ 
  23 & 71287\_48 & lipolysis-stimulated lipoprotein receptor \\ 
  24 & 74776\_68 & baculoviral iap repeat-containing protein 6 isoform x10 \\ 
  25 & 79151\_39 & guanine nucleotide-binding protein g g g subunit beta-1-like isoform x1 \\ 
  26 & 81519\_68 & nuclear receptor corepressor 1 isoform x3 \\ 
  27 & 89719\_68 & unnamed protein product, partial \\ 
   \hline
\end{tabular}
\end{table}

>>>>

To emphasize the utility of RDA, we compared to pcadapt from which the idea of using multivariate method for genome scan is based. On the simulated dataset, we retain for both methods $K=5$ axis to compute Mahalanobis distances  as it seems to explain the main amount of variance in the data using scatter plots. To control for false positive, we used the same qvalue threshold (i.e. <= 10\%).

@@

	A lattice of 8x8 populations is simulated (i.e. 64 populations in total). Each population is initialized with 200 diploid individual with random genotypes. Migration is set to 0.1 so that population structure must be very smooth and genetic differentiation must show an isolation by distance pattern over the 64 populations. This is where pcadapt is best designed for. Loci are biallelic (0 or 1) like SNPs. Allele frequency of the whole population is initialized at 0.5. 1000 loci are defined. They are separated in 200 chuncks of 5 SNPs in physical linkage with recombination rate between adjacent loci fixed at 0.1. 3 different Traits are coded by a group of 10 different loci. The first trait is coded by loci 1, 11, 21, ..., 91. The trait value is simply the sum of genotype value and therefore can take value between 0 an 20. For the sake of realism, we add to each trait a random noise (non heritable variation) drawn from a normal distribution $N(0,2)$. The second trait is coded by loci 101, 111, ..., 191 and the third is coded by loci 201, 211, ..., 291. Each trait is therefore coded by free recombining SNP loci. In other words, there are 30 coding SNPs among 1000. Selection can have an effect on linked loci, for instance, loci 2, 3, 4 and 5 can be impacted by selection on locus 1. However, recombination is high enough (0.1) to expect a limited linkage effect. We have defined 10 different environmental variables. The first one determines the selective pressure on trait 1, the second one on trait 2 and the third one on trait 3. The first environment variable is a quadratic gradient coded by function  $env1 = -(\cos(\theta)*(i-3.5))^2 -(sin(\theta)*(j-3.5))^2 + 18, \theta = \pi/2$, i and j being the population indicator on the 8x8 lattice. The second one is a linear plan gradient coded by function $env2 = h*\cos(\theta)*(i-1) + h*\sin(\theta)*(j-1) + k$ with $h=2$, $\theta = \pi/4$ and $k=3$. The third environment variable simulates a coarse environment with value $env3 = 2$ for all populations except population (i,j) = {(2,2), (2,3), (3,2), (3,3), (6,2), (6,3), (7,2), (7,3), (2,6), (2,7), (3,6), (3,7), (6,6), (6,7), (7,6), (7,7)} for which $env3 = 18$. Env4, env5 and env6 have exactly the same equation than env1, env2 and env3 respectively. The remaining 4 environment variable are similar to env2 but with different value of h and $\theta$. Env7 has $h=2$, $\theta = 0$ and $k=3$. Env8 has $h=2$, $\theta = \pi/4$ and $k=0$. Env9 has $h=1$, $\theta = \pi/4$ and $k=4$. Env10 has $h=0.5$, $\theta = \pi/4$ and $k=8$. Graphical representation of mean environmental value for environment 1, 2 and 3 is given in Fig. \ref{simulatedenvir}. Environment 4, 5 and 6 have respectively the same mean value spatial distribution. For a graphical representation of environment 7 to 10, see supplementary material. Environmental equation gives a mean value of the environment variable. To avoid colinearity between environments variable, we added noise by drawing an environment value within a normal distribution $N(\mu=env, \sigma=1)$. Fitness for each trait is set to be $-e^{((x-env)^2/(2*\omega^2))}$, $x$ being the quantitative trait value, $env$ the environmental value and $\omega$ is defining selection strength and has been set to 20 which in our experience seems sufficient for loci to be often detected. To get the overall fitness for a given individual, fitness associated to each trait are multiplied. Fitness are relative and selection arises on parents and determine their number of offsprings. Simulations are made across 500 generations. At the end of simulation, we sample 10 individuals per population. Therefore, we have a sample of 640 individuals with 1000 SNP-like loci.

\subsection{Real dataset}

The \textit{Populus trichocarpa} dataset is a sample of 424 individuals genotyped on 33,070 SNPs from 25 drainages (i.e., topographic units separated by watershed barriers) \citep{Geraldes2014}. Genotyping of each accession was performed as described in \citep{Geraldes2013} using a 34K Populus SNP array targeting 34,131 SNPs mostly within (plus 2kb upstream and downstream) 3543 genes. Details of SNP and gene selection can be found in \citep{Geraldes2013}. 21 climatic variables are available on each sampling site (details in Table S1). For the RDA and pcadapt analysis, we retained respectively $K=6$ and $K=10$ axis as they seem to explain the majority of variance in the data. From the 33,079 SNPs, we removed the SNPs with missing values, leaving us 19,336 SNPs.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[height=0.21\textheight]{figures/sim105_pcadapt.eps}
\end{center}
\caption{Manhattan plot of the result of pcadapt on a simulated data set.}%
\label{fig:pcadapt}%
\end{figure}

\begin{figure}[t]
\begin{center}
\includegraphics[height=0.21\textheight]{figures/sim105_capscale.eps}
\end{center}
\caption{Manhattan plot of the result of genome scan using RDA on a simulated data set.}%
\label{fig:rda}%
\end{figure}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[height=0.4\textheight]{figures/rdaA.eps}
\end{center}
\caption{RDA on the adaptively enriched genetic space. We discarded the individual points for readability. Dots represents outliers SNPs.  $R^2$ of envir1 with the first, second and third axis is (0.172\%, 77.6\%, 17.6\%), envir2 is (94.5\%, 0.560\%, 0.236\%) and envir3 is (0.189\%, 5.34\%, 90.6\%)}%
\label{fig:rdaA}%
\end{figure*}

@@

\subsection{\textit{Populus trichocarpa}}

Analysis of \textit{P. trichocarpa} with fdr of $0.1$ leads to a list of 105 outliers for RDA and 122 for pcadapt. Interstingly, both methods have 52 outliers in common so that 53 and 70 SNPs are outliers specific to respectively RDA and pcadapt. When we compared RDA genome scan results with the outliers found by \citet{Geraldes2014} based on Fdist, bayescan and bayenv methods, we found that substential overlap between them. However, RDA based genome scan found 35 SNPs that no other methods (including pcadapt) had detected as outlier (see Table SXX).

@@

Fig \ref{fig:pcadapt} shows that pcadapt approach works well when the environmental gradient and the selective pressures are correlated to the  geographical pattern of isolation by distance. Whereas when the environment is a coarse environment (QTL3) it fails dramatically. Indeed, we can hypothesize that the PCA ordination in this case is not able to orientiate the genetic space differentiation into the direction of environment 3 therefore leaving no chance to detect any outliers on the QTL influenced by this environmental variable. Fig \ref{fig:rda} shows that RDA has a much better behavior than pcadapt by taking advantage of using informations of environmental local conditions. It can be attributed to a better alignment between the genetic space and the environmental variable improving the power to detect true positives.

Both methods have a good control of false discovery rate although slightly better for RDA ($12.0 \times 10^{-2}$ for pcadapt and $10.6 \times 10^{-2}$ for RDA). Results summarized on Fig \ref{fig:performance} is confirming that overall RDA shows better performance at detecting true outliers since it succeeds to detect quite often QTL1 and QTL3 SNPs. It seems however less efficient at detecting QTL3 outliers but this might be due to the fact that local adaptation on a coarse environment is more difficult  that adaptation on a smooth environmental gradient as environment 1 and 2. These simulations plead in favor of using constrained ordination method instead of PCA when non genetic data such as environmental variable are available in order to orientate the axis in the direction of informative gradients.

@@

From the analysis of \textit{Popular trichocarpa}, we picked up some genes that can be interpreted regarding to the environmental variable. For instance, among the top SNPs, some of them are located within or nearby, POPTR_0015s00640 implicated in cellular response to cold, POPTR_0015s00440 involved in the regulation of circadian clock in plants. One outlier, POPTR_0017s06840 that is not detected by pcadapt is involved in amylase catabolism, an activity that is suspected to depend of climatic conditions. Interestingly, Table S2 shows that a substantial amount of RDA genome scan outliers (35) are not detected by any other genome scan method including pcadapt. For instance, four SNPs are assigned to POPTR_0007s04340, a dehydration-responsive protein-related which is similar to similar to early-responsive to dehydration stress ERD3 protein. However, its function remains mainly unknown. 
When performing an RDA analysis of adaptively enriched genetic space (Fig. S2), we noticed that first and second axis of RDA explains a large amount of constrained variance and can be used to identify important environmental gradient. RDA axis 1 This was of course expected since the ordination is constrained by environmental axis. One can identify an outlier to a selective gradient which was used in the inference process whereas outliers picked up by PCA based method have to be interpreted a posteriori and this can be more difficult to justify. It is however noticeable that a large proportion of outliers are shared between PCA and RDA based methods which highlight the fact that both approach are very similar. The lack of interpretability of PCA method is compensated by the fact that it is a blind method in regard to environmental data so it can pick up genes that we will miss using RDA because some crucial environmental data are lacking. 
In conclusion, this paper shows that both methods are complementary and should be used simultaneously on the same dataset. RDA will be more easily interpretable in terms of mechanism and should pick up more genes when some important selective gradient are identified but PCA might be able to pick up outliers where no environmental data are available. However this latter method will miss outliers when selective gradient are poorly correlated to the population structure (i.e. geographical distance between populations and individuals). It is therefore crucial when performing landscape genomics to clearly characterize the environmental condition by measuring important variable that are suspected to put selective pressures on the species of interest. This can be done through RDA as it has been validated in our simulations and in the real dataset on Chinook Salmon. 
A myriad of constrained ordination method exist that could be used and allows to control for confounding variable (i.e. partial RDA) for instance altitude, latitude and longitude or allow to perform none linear regression between genetic and environmental data (i.e. LVM - Latent Variable Model). This is still to be tested and explored using both simulation and real data set validation.

### 2017-03-13 suppmat/suppmat.tex
update

>>>>

\begin{figure*}[t]
\begin{center}
\includegraphics[height=0.6\textheight]{figures/poplar_rda.eps}
\end{center}
\caption{RDA analysis on \textit{Poplar trichocarpa} adaptive enriched genetic space. The first two axis RDA1 and RDA2 explain 37.8 and 14.0\% of variance predicted by environmental variable (22.7\% of constrained variance). Specific and shared outliers between PCA based method and RDA based method are differentially colored.}%
\label{fig:poplarrda}%
\end{figure*}

\begin{figure*}[t]
\begin{center}
\includegraphics[height=0.6\textheight]{figures/corrplot.eps}
\end{center}
\caption{biplot of RDA analysis. Color and circle diameter are proportional to the amount of linkage between RDA axis and climatic variable.}%
\label{fig:corrplot}%
\end{figure*}
